---
title: "01_data_manipulation2"
author: "Madeline Sharp"
date: "2026-09-29"
output: github_document
---

This file is for doing data manipulation.

``` {r}
library(tidyverse)
```

Load the litters and pups datasets.

```{r}
litters_df =
  read_csv(
    "Data/FAS_litters.csv",
    na = c("","NA", ".")
  )

litters_df = janitor::clean_names(litters_df)

pups_df =
  read_csv(
    "Data/FAS_pups.csv",
    #skip = 3,
    na = c("","NA", ".")
  )

pups_df = janitor::clean_names(pups_df)
```


## `select` some variables

```{r}
select(litters_df, group, litter_number)

select(litters_df, group:gd_of_birth)

select(litters_df, -group)

select(litters_df, starts_with("gd"))

select(litters_df, starts_with("gd"), group)

select(litters_df, GROUP = group, everything())

select(litters_df, LITTER = litter_number, everything())

select(litters_df, -starts_with("gd"))
```

a couple of alternatives

```{r}
# shorthand version of `select(litters_df, LITTER = litter_number, everything())`
rename(litters_df, GROUP = group)

# Jeff prefers subject ID is 1st variable. Put things to the front that are the highest levels of hierarchy in your data. `relocate()` puts object (cols) in a different order
relocate(litters_df, litter_number, gd0_weight)
```

In the pups data, select litter number, sex and PD ears.

```{r}
#select(pups_df, litter_number, sex, pd_ears)

select(pups_df, x1, x1_male_2_female, x3)
```

## 'filter` some rows

```{r}
filter(litters_df, gd_of_birth == 19)

filter(litters_df, pups_born_alive <= 5)

filter(litters_df, pups_born_alive != 6)

filter(litters_df, group %in% c("Con7", "Con8"))

filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive < 7)

filter(litters_df, pups_born_alive < 4 | pups_born_alive > 8)

drop_na(litters_df, gd0_weight)
```


In the pups dataset, keep only rows where PD walk is less than 11 and sex is 2

```{r}
filter(pups_df, x3 < 11, x1_male_2_female == 2)
```


## `mutate` to change some columns

let's add and change soem columns

```{r}
# you can create and modify variables right when you've made them
mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight,
  wt_gain_squared = wt_gain^2,
  group = str_to_lower(group))
```

In the pups dataset, create a variable that subtracts 7 from PD walk

```{r}
#ERROR HELP!!
mutate(pups_df, x6_minus_7 = as.numeric(x6) - 7)
```

## `arrange` the data

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, desc(pups_born_alive))
```

## Do multiple steps...

This is bad:
```{r}
# never do this. This is bad
litters_df = read_csv("Data/FAS_litters.csv", na = c("", "NA", "."))
litters_rename_df = janitor::clean_names(litters_df)
litters_with_gd_df = select(litters_rename_df, group, starts_with("gd"))
litters_no_na_df = drop_na(litters_with_gd_df)
litters_wt_gain_df = mutate(litters_no_na_df, wt_gain = gd18_weight - gd0_weight)
litters_wt_gain_df
```

This is worse:

```{r}
select(
  janitor::clean_names(
    read_csv("data/FAS_litters.csv", na = c("", "NA", "."))), 
  group, starts_with("gd"))
```

This is good
```{r}
litters_df =
  read_csv("Data/FAS_litters.csv", na = c("","NA",".")) |>
  janitor::clean_names() |>
  select(group, starts_with("gd")) |>
  drop_na() |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )
```

Load pups, clean names, dop missing, keep group and pd variables, add pd walk ~7
```{r}
pups_df =
  read_csv("Data/FAS_pups.csv", skip = 3, na = c("","NA",".")) |>
  janitor::clean_names() |>
  drop_na() |>
  select(litter_number, starts_with("pd")) |>
  mutate(
    pd_walk_minus_7 = pd_walk - 7
  )
```

sometimes the first thing isn't as accurate

```{r}
litters_df |>
  lm(gd18_weight ~ gd0_weight, data = _)
```

```{r}
litters_no_na_df |> 
  filter(group %in% c("Con7", "Con8")) |> 
  lm(gd18_weight ~ gd0_weight, data = _)
```

```{r}
litters_df |>
  select(group)

litters_df |>
  filter(group %in% c("con7","con8")) |>
  pull(group)
```

never do `$`

```{r}
litters_df$group
```


pipes and viewing

```{r}
litters_df |>
  filter(group %in% c("con7","con8"))

#don't do this in R Markdown
#View(litters_df)
```